Extended Data

### 2022-03-11 doc/manuscript/Dow_manuscript_Nature.Rmd
minor edits

To understand how annual growth increments have responded to spring temperatures at the centennial scale, we analyzed tree-ring chronologies of `r n_chronologies_SCBI` species at SCBI[@helcoski_growing_2019] and `r n_chronologies_HF` species at Harvard Forest (`r EDT_species`), along with an additional `r n_chronologies-n_chronologies_SCBI-n_chronologies_HF` chronologies from `r n_cores_sites-2` sites (Fig. `r tree_rings`; `r EDF_map`; `r EDT_tree_ring_summary`)[@dorangeville_drought_2018].
In total, our analysis included `r n_chronologies` chronologies representing `r n_cores_species` broadleaf species at `r n_cores_sites` sites distributed from Alabama (Lat = 34.35) to Michigan (Lat = 45.56) and spanning a 15 $^\circ$C range in April $T_{max}$.
Across all chronologies, the standardized ring-width index (RWI) was significantly (at p $\le$ 0.05) positively correlated with April $T_{max}$ for only 2% of chronologies: 1 of 142 ring porous and 4 of 66 diffuse porous species-site combinations (`r EDT_tree_ring_summary`). 
There were even fewer significant positive correlations with March and May $T_{max}$: 0% and `r round(2/(142+66),2)*100`%, respectively (`r EDT_tree_ring_summary`).
In contrast, RWI was frequently negatively correlated with $T_{max}$ during peak growing season months (May-August), with significant correlations for 52% (May: 45/141, Jun: 107/141, Jul: 91/141, Aug: 53/141) and 46% (May: 10/66, Jun: 52/66, Jul: 36/66, Aug: 23/66) of species-site-month combinations for ring porous and diffuse porous species, respectively.
$T_{min}$ generally exhibited weaker relationships to annual growth than $T_{max}$, with few significant correlations between spring $T_{min}$ and RWI (`r EDF_tree_rings_TMIN`).

@@

<!--Currently 1005. Goal: 700-800-->

Together, our results demonstrate that warmer spring temperatures in the temperate deciduous forests of eastern North America advance the timing of stem growth but have little effect on total annual growth (Figs. `r schematic_summary`- `r tree_rings`). 
The observed advance in stem growth under warmer springs parallels phenological advances observed for canopy foliage (Fig. `r DOY_timing`a-b)[@keenan_net_2014;@finzi_carbon_2020;@stuble_plant_2021] and _NEE_[@keenan_net_2014;@finzi_carbon_2020]. 
However, inconsistent with the concept that an earlier start to growth would increase annual woody productivity, we demonstrate that warmer springs either hasten the deceleration of stem expansion or otherwise fail to translate extended growing seasons into biologically significant increases in stem growth (Fig. `r schematic_summary`), and thereby have negligible effect on total annual growth for most species and locations (Fig. r tree_rings`).
Our observations suggest that the deceleration of stem expansion (starting around $DOY_{50}$), which occurs mid-summer near the time of peak canopy greenness (`r EDF_growth_curves_all`)[@finzi_carbon_2020;@etzold_number_2022], is likely driven by cues other than photosynthate limitation, such as water stress, daylength, or sink limitation[@zweifel_why_2021;@etzold_number_2022;@zani_increased_2020;@keenan_timing_2015;@fu_variation_2014;@delpierre_wood_2016;@delpierre_temperate_2016].
Our tree-ring analysis (Fig. `r tree_rings`) demonstrates that the primary effect of warming temperatures on annual ring-width is not an augmentation through an earlier start to growth, but rather a reduction associated to drought stress during the peak growing season[@dorangeville_drought_2018;@etzold_number_2022]. 
Warm springs may also amplify summer drought stress in some times and places, effectively canceling out any positive effects of an extended growing period[@buermann_widespread_2018;@zhang_drought_2021;@lian_summer_2022].

@@

- stem growth is routinely used to estimate woody productivity (but wood density partially decouples them) [@delpierre_wood_2016 and refs therein]
- positive effects of warm springs can be compensated by negative effects of accumulation of seasonal water deficits[@buermann_widespread_2018;@lian_summer_2022]
- our study provides further evidence of sink limitation to growth
- we do not know where the extra C is going. If we look at instances where growth is more strongly limited than GPP, we see that NSCs can build up[@refs], and this build up can feed back to reduce photosynthesis[@refs]. We can also look to studies documenting enhanced GPP with no enhancement of growth. In a mature eucalypt forest exposed to elevated CO2, enhanced GPP did not result in higher growth, but rather was released back to the atmosphere on relatively short time scales through enchanced respiration[@jiang_fate_2020]
~~~

>>>>

To understand how annual growth increments have responded to spring temperatures at the centennial scale, we analyzed tree-ring chronologies of `r n_chronologies_SCBI` species at SCBI[@helcoski_growing_2019] and `r n_chronologies_HF` species at Harvard Forest (`r EDT_species`), along with an additional `r n_chronologies-n_chronologies_SCBI-n_chronologies_HF` chronologies from `r n_cores_sites-2` sites (Fig. `r tree_rings`; `r EDF_map`; `r EDT_tree_ring_summary`)[@dorangeville_drought_2018].
In total, our analysis included `r n_chronologies` chronologies representing `r n_cores_species` broadleaf species at `r n_cores_sites` sites distributed from Alabama (34.35$^\circ$N latitude) to Michigan (45.56$^\circ$N latitude) and spanning a 15 $^\circ$C range in April $T_{max}$.
Across all chronologies, the standardized ring-width index (RWI) was significantly (at p $\le$ 0.05) positively correlated with April $T_{max}$ for only 2% of chronologies: 1 of 142 ring porous and 4 of 66 diffuse porous species-site combinations (`r EDT_tree_ring_summary`). 
There were even fewer significant positive correlations with March and May $T_{max}$: 0% and `r round(2/(142+66),2)*100`%, respectively (`r EDT_tree_ring_summary`).
In contrast, RWI was frequently negatively correlated with $T_{max}$ during peak growing season months (May-August), with significant correlations for 52% (May: 45/141, Jun: 107/141, Jul: 91/141, Aug: 53/141) and 46% (May: 10/66, Jun: 52/66, Jul: 36/66, Aug: 23/66) of species-site-month combinations for ring porous and diffuse porous species, respectively.
$T_{min}$ generally exhibited weaker relationships to annual growth than $T_{max}$, with few significant correlations between spring $T_{min}$ and RWI (`r EDF_tree_rings_TMIN`).

@@

<!--1. summary-->
Together, our results demonstrate that warmer spring temperatures in the temperate deciduous forests of eastern North America advance the timing of stem growth but have little effect on total annual growth (Figs. `r schematic_summary`- `r tree_rings`). 
The observed advance in stem growth under warmer springs parallels phenological advances observed for canopy foliage (Fig. `r DOY_timing`a-b)[@keenan_net_2014;@finzi_carbon_2020;@stuble_plant_2021] and _NEE_[@keenan_net_2014;@finzi_carbon_2020]. 
However, inconsistent with the concept that an earlier start to growth would increase $ANPP_{woody}$, we demonstrate that warmer springs either hasten the deceleration of stem expansion or otherwise fail to translate extended growing seasons into biologically significant increases in stem growth (Fig. `r schematic_summary`), and thereby have negligible effect on total annual growth for most species and locations (Fig. `r tree_rings`).
Our observations suggest that the deceleration of stem expansion (starting around $DOY_{50}$), which occurs mid-summer near the time of peak canopy greenness (`r EDF_growth_curves_all`)[@finzi_carbon_2020;@etzold_number_2022], is likely driven by cues other than photosynthate limitation, such as water stress, daylength, or sink limitation[@zweifel_why_2021;@etzold_number_2022;@zani_increased_2020;@keenan_timing_2015;@fu_variation_2014;@delpierre_wood_2016;@delpierre_temperate_2016].
Our tree-ring analysis (Fig. `r tree_rings`) demonstrates that the primary effect of warming temperatures on annual ring-width is not an augmentation through an earlier start to growth, but rather a reduction associated to drought stress during the peak growing season[@dorangeville_drought_2018;@etzold_number_2022].

Warm springs may also amplify summer drought stress in some times and places, effectively canceling out any positive effects of an extended growing period[@buermann_widespread_2018;@zhang_drought_2021;@lian_summer_2022].

@@

- stem growth is routinely used to estimate woody productivity (but wood density partially decouples them) [@delpierre_wood_2016 and refs therein]
- positive effects of warm springs can be compensated by negative effects of accumulation of seasonal water deficits[@buermann_widespread_2018;@lian_summer_2022]
- our study adds to a growing body of literature demonstrating a sink limitation of stem growth and $ANPP_{woody}$ [@refs]. Specifically, [explain]. The implication is that while elements of global change such as warming spring temperatures or elevated CO~2~[@refs] may enhance GPP, woody growth remains constrained.
- we do not know where the extra C is going. If we look at instances where growth is more strongly limited than GPP, we see that NSCs can build up[@refs], and this build up can feed back to reduce photosynthesis[@refs]. We can also look to studies documenting enhanced GPP with no enhancement of growth. In a mature eucalypt forest exposed to elevated CO2, enhanced GPP did not result in higher growth, but rather was released back to the atmosphere on relatively short time scales through enchanced respiration[@jiang_fate_2020]
~~~
